DEG_RNA_summary

### 2025-03-06 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd
cleanup analyses code

dir_out <- "/home/common/data/output/projects/ENDO/E044/A031/"
setwd(dir_out)
```

Determine inflection point based on unfiltered DEGs
```{r}
##Combine muscat output into one dataframe
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
RNAlistRDS <- lapply(PATHS, readRDS)
names(RNAlistRDS) <- print(PATHS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))

@@

summarize DEGs into df and plot
```{r RNA assay}
#RNA assay
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
print(PATHS)
RNAlistRDS <- lapply(PATHS, readRDS)
names(RNAlistRDS) <- print(PATHS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))
RNAlistRDS005 <- lapply(RNAlistRDS, function(x) dplyr::filter(x, p_adj.loc < 0.05))
RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))
#Check for cluster_ids that have or don't have DEGs with p_adj.loc < 0.05
intersect(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))
setdiff(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))
#Add columns stage, clustering, annotation_reference
library(stringr)
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(stage = case_when(
    str_detect(rownames(RNAlistRDS005), "_severe") ~ "severe",
    str_detect(rownames(RNAlistRDS005), "_mild") ~ "mild",
    TRUE ~ "all"
  ))
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(clustering = case_when(
    str_detect(rownames(RNAlistRDS005), "major") ~ "major",
    str_detect(rownames(RNAlistRDS005), "united") ~ "united",
    TRUE ~ "RefinedCluster"
  ))
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(annotation_reference = case_when(
    str_detect(rownames(RNAlistRDS005), "Tan") ~ "Tan",
    str_detect(rownames(RNAlistRDS005), "Fonseca") ~ "Fonseca"
  ))

@@

write.csv(RNAlistRDS005,"/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
#RNAlistRDS005 <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv", row.names = 1)

gProfiler based on RNA assay and per annotation with gene expression >18.53202
```{r}
##function to run gProfiler and save output
gProfiler_FUN <- function(x,ANALYSIS,GRADE,ORDERED_QUERY,ASSAY,dir_out) {
#gProfiler
custom_bg <- unique(x$gene)
query <- x %>%
  dplyr::filter(AveExpr > 18.53202) %>%
  dplyr::filter(p_adj.loc < 0.05) %>%
  dplyr::arrange(p_adj.loc) #arrange by lowest p_adj.loc
if(nrow(query) > 0){
query <- distinct(query, gene, .keep_all= TRUE) #remove duplicates of gene with higher p_adj.loc
query <- query$gene
gostres <<- gost(query = query, 
                organism = "hsapiens", 
                ordered_query = ORDERED_QUERY, #can use TRUE, p_adj.loc has been arranged 
                multi_query = FALSE, 
                significant = TRUE, #only statistically significant results should be returned
                exclude_iea = FALSE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE,
                user_threshold = 0.05, 
                correction_method = "g_SCS", #g_SCS is default
                domain_scope = "annotated", #if custom_bg is set, I believe this shouldn't matter too much
                custom_bg = custom_bg, #custom background, use!
                numeric_ns = "", 
                sources = NULL, #by default all sources are analyzed
                as_short_link = FALSE)
if(length(gostres) > 0){
results <- gostres$result
results$cluster_id <- levels(as.factor(x$cluster_id))
results$analysis <- print(ANALYSIS)
results$assay <- print(ASSAY)
results$grade <- print(GRADE) 
results$ordered_query <- print(ORDERED_QUERY)
results$parents <- NULL
write.csv(results,paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
print(paste0("Saved gotres as gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
plot <- gostplot(gostres, capped = TRUE, interactive = FALSE) + ggtitle(paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY)))
ggsave(plot = plot, filename = paste0(dir_out,"gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".png"), width = 7, height = 5)
print("Saved plot")
}
}
}
##function to run gProfiler_FUN on different grades, analysis and assay
RUN_gProfier <- function(GRADE, ANALYSIS,ASSAY,SPECIFICATION){
dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ANALYSIS,"/",ASSAY,"assay_",SPECIFICATION,"/")
dir.create(dir_out, recursive = TRUE)
setwd(dir_out)
#Read paths to DEG output .rds files
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)

>>>>

dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/05b_Functional_enrichment_analysis/"
setwd(dir_out)

# define menstrual cycle phase 
MenstrualCyclePhase = ""

# define annotation level 
Annotation = ""

```
Determine inflection point based on unfiltered DEGs
```{r}

##Combine muscat output into one dataframe
PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/",MenstrualCyclePhase,"_", Annotation), pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
RNAlistRDS <- lapply(PATHS, readRDS)
names(RNAlistRDS) <- print(PATHS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))

@@

```
summarize DEGs into df and plot
```{r RNA assay}
#RNA assay
PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/",MenstrualCyclePhase,"_", Annotation), pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
print(PATHS)
RNAlistRDS <- lapply(PATHS, readRDS)
names(RNAlistRDS) <- print(PATHS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))
RNAlistRDS005 <- lapply(RNAlistRDS, function(x) dplyr::filter(x, p_adj.loc < 0.05))
RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))

#Check for cluster_ids that have or don't have DEGs with p_adj.loc < 0.05
intersect(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))
setdiff(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))

#Add columns stage, clustering, annotation_reference
library(stringr)
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(stage = case_when(
    str_detect(rownames(RNAlistRDS005), "_severe") ~ "severe",
    str_detect(rownames(RNAlistRDS005), "_mild") ~ "mild",
    TRUE ~ "all"
  ))
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(clustering = case_when(
    str_detect(rownames(RNAlistRDS005), "major") ~ "major",
    str_detect(rownames(RNAlistRDS005), "united") ~ "united",
    TRUE ~ "RefinedCluster"
  ))
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(annotation_reference = case_when(
    str_detect(rownames(RNAlistRDS005), "Tan") ~ "Tan",
    str_detect(rownames(RNAlistRDS005), "Fonseca") ~ "Fonseca"
  ))

@@

write.csv(RNAlistRDS005,file.path(dir_out,"RNAlistRDS005_unmerged.csv"))

```
gProfiler based on RNA assay and per annotation with gene expression >18.53202
```{r}
##function to run gProfiler and save output
gProfiler_FUN <- function(x,ANALYSIS,GRADE,ORDERED_QUERY,ASSAY,dir_out) {
#gProfiler
custom_bg <- unique(x$gene)
query <- x %>%
  dplyr::filter(AveExpr > 18.53202) %>%
  dplyr::filter(p_adj.loc < 0.05) %>%
  dplyr::arrange(p_adj.loc) #arrange by lowest p_adj.loc
if(nrow(query) > 0){
query <- distinct(query, gene, .keep_all= TRUE) #remove duplicates of gene with higher p_adj.loc
query <- query$gene
gostres <<- gost(query = query, 
                organism = "hsapiens", 
                ordered_query = ORDERED_QUERY, #can use TRUE, p_adj.loc has been arranged 
                multi_query = FALSE, 
                significant = TRUE, #only statistically significant results should be returned
                exclude_iea = FALSE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE,
                user_threshold = 0.05, 
                correction_method = "g_SCS", #g_SCS is default
                domain_scope = "annotated", #if custom_bg is set, I believe this shouldn't matter too much
                custom_bg = custom_bg, #custom background, use!
                numeric_ns = "", 
                sources = NULL, #by default all sources are analyzed
                as_short_link = FALSE)
if(length(gostres) > 0){
results <- gostres$result
results$cluster_id <- levels(as.factor(x$cluster_id))
results$analysis <- print(ANALYSIS)
results$assay <- print(ASSAY)
results$grade <- print(GRADE) 
results$ordered_query <- print(ORDERED_QUERY)
results$parents <- NULL
write.csv(results,paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
print(paste0("Saved gotres as gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
plot <- gostplot(gostres, capped = TRUE, interactive = FALSE) + ggtitle(paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY)))
ggsave(plot = plot, filename = paste0(dir_out,"gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".png"), width = 7, height = 5)
print("Saved plot")
}
}
}
##function to run gProfiler_FUN on different grades, analysis and assay
RUN_gProfier <- function(GRADE, ANALYSIS,ASSAY,SPECIFICATION){
dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ANALYSIS,"/",ASSAY,"assay_",SPECIFICATION,"/")
dir.create(dir_out, recursive = TRUE)
setwd(dir_out)
#Read paths to DEG output .rds files
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
